---
title: "Data607_Final_Project"
author: "Mahmud Hasan Al Raji"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Gun violence, especially mass shootings is now a disease in USA. Therefore, I think it is an interesting topic to understand and analyze if there is a correlation between unemployment rate and mass shootings in the USA. So, i set the research question for this project: "Does unemployment rate have an effect on mass shootings in US states?" For this analysis, I collected the unemployment rate data for the states from The U.S. Bureau of Labor Statistics for the year between 2017 and 2021 by using web scrapping from their web page. I also collected  mass shooting casualty data as CSV files from the Gun Violence Archive and stored them on my github.Then, I will clean up and transform the data as required to conduct some exploratory and statistical analysis (linear regression analysis) to see if there is a correlation between the stated variables.   

# Load libraries
```{r message=FALSE}
library(dplyr)
library(rvest)
library(ggplot2)
```

# Getting unemployment rate data for states for the year between 2017 and 2021 through web srcapping from Us Bureau of Labor Statistics website and creating data frames 

```{r }
# Uemployment rate data of 2021 annual averages
unemp2021_url<-read_html("https://www.bls.gov/lau/lastrk21.htm")
unemp_data_2021<-unemp2021_url %>% html_nodes(xpath='//*[@id="lastrk21"]/tbody') %>% html_table()
unemp2021_df<-as.data.frame(unemp_data_2021)
glimpse(unemp_data_2021)
head(unemp2021_df)

# Unemployment rate data of 2020 annual averages
unemp2020_url<-read_html("https://www.bls.gov/lau/lastrk20.htm")
unemp_data_2020<-unemp2020_url %>% html_nodes(xpath='//*[@id="lastrk20"]/tbody') %>% html_table()
unemp2020_df<-as.data.frame(unemp_data_2020)
head(unemp2020_df)

# Unemployment rate data of 2019 annual averages
unemp2019_url<-read_html("https://www.bls.gov/lau/lastrk19.htm")
unemp_data_2019<-unemp2019_url %>% html_nodes(xpath='//*[@id="lastrk19"]/tbody') %>% html_table()
unemp2019_df<-as.data.frame(unemp_data_2019)
head(unemp2019_df)

# Unemployment rate data of 2018 annual averages
unemp2018_url<-read_html("https://www.bls.gov/lau/lastrk18.htm")
unemp_data_2018<-unemp2018_url %>% html_nodes(xpath='//*[@id="lastrk18"]/tbody') %>% html_table()
unemp2018_df<-as.data.frame(unemp_data_2018)
head(unemp2018_df)

# Unemployment rate data of 2017 annual average
unemp2017_url<-read_html("https://www.bls.gov/lau/lastrk17.htm")
unemp_data_2017<-unemp2017_url %>% html_nodes(xpath='//*[@id="lastrk17"]/tbody') %>% html_table()
unemp2017_df<-as.data.frame(unemp_data_2017)
head(unemp2017_df)
```

# Unemployment rate data tidying and transformation
```{r }
# Remove irrelevant rows and columns from data frames
unemp2021_df<-unemp2021_df %>% slice(-c(1,2))
unemp2021_df<-subset(unemp2021_df, select=-X3)

unemp2020_df<-unemp2020_df %>% slice(-c(1,2))
unemp2020_df<-subset(unemp2020_df, select=-X3)

unemp2019_df<-unemp2019_df %>% slice(-c(1,2))
unemp2019_df<-subset(unemp2019_df, select=-X3)

unemp2018_df<-unemp2018_df %>% slice(-c(1,2))
unemp2018_df<-subset(unemp2018_df, select=-X3)

unemp2017_df<-unemp2017_df %>% slice(-c(1,2))
unemp2017_df<-subset(unemp2017_df, select=-X3)

# Renaming columns
unemp2021_df<-unemp2021_df %>% rename(state=X1,unemployment_rate_2021=X2)
head(unemp2021_df)

unemp2020_df<-unemp2020_df %>% rename(state=X1,unemployment_rate_2020=X2)
head(unemp2020_df)

unemp2019_df<-unemp2019_df %>% rename(state=X1,unemployment_rate_2019=X2)
head(unemp2019_df)

unemp2018_df<-unemp2018_df %>% rename(state=X1,unemployment_rate_2018=X2)
head(unemp2018_df)

unemp2017_df<-unemp2017_df %>% rename(state=X1,unemployment_rate_2017=X2)
head(unemp2017_df)

# Combining data frames with one column name in common
unemp_df<-full_join(unemp2021_df,unemp2020_df,by="state") %>% 
          full_join(unemp2019_df,by="state") %>% 
          full_join(unemp2018_df,by="state") %>% 
          full_join(unemp2017_df,by="state")
head(unemp_df)

# Convert to lowercase in data frames
unemp_df$state<-tolower(unemp_df$state)
head(unemp_df)

# Sorting data frame by alphabetic order
unemp_df <- unemp_df[order(unemp_df$state),]
head(unemp_df)

# Adding average unemployment rate column
unemp_df<-unemp_df %>%
  rowwise() %>%
  mutate(avg_unemp_rate = mean(c_across(unemployment_rate_2021:unemployment_rate_2017))) %>% select(state,avg_unemp_rate)
glimpse(unemp_df)
head(unemp_df)
```

# Getting mass shooting casualty data 
```{r }
df_2017<-read.csv("https://raw.githubusercontent.com/Raji030/data607_mass_shooting/main/2017.csv")
df_2018<-read.csv("https://raw.githubusercontent.com/Raji030/data607_mass_shooting/main/2018.csv")
df_2019<-read.csv("https://raw.githubusercontent.com/Raji030/data607_mass_shooting/main/2019.csv")
df4<-read.csv("https://raw.githubusercontent.com/Raji030/data607_mass_shooting/main/19-22.csv")
```
# Mass shooting casualty data tidying and transformation
```{r }
# Subset data frame after columns renaming and adding new column
df_2017<-df_2017 %>% rename(state=State,killed=X..Killed,injured=X..Injured)
df_2017<-df_2017 %>% mutate(total_casualty_2017=killed+injured) %>% select(state,total_casualty_2017)

df_2018<-df_2018 %>% rename(state=State,killed=X..Killed,injured=X..Injured)
df_2018<-df_2018 %>% mutate(total_casualty_2018=killed+injured) %>% select(state,total_casualty_2018)

df_2019<-df_2019 %>% rename(state=State,killed=X..Killed,injured=X..Injured)
df_2019<-df_2019 %>% mutate(total_casualty_2019=killed+injured) %>% select(state,total_casualty_2019)

df_2020 <- df4[c(1310:1919), ]
df_2020<-df_2020 %>% rename(state=State,killed=X..Killed,injured=X..Injured)
df_2020<-df_2020%>% mutate(total_casualty_2020=killed+injured) %>% select(state,total_casualty_2020)

df_2021<-df4[c(620:1309),]
df_2021<-df_2021%>% rename(state=State,killed=X..Killed,injured=X..Injured)
df_2021<-df_2021%>% mutate(total_casualty_2021=killed+injured) %>% select(state,total_casualty_2021)

#Merging rows with the same name in state column
df_2017<-df_2017 %>% group_by(state) %>% summarise_each(funs(sum))
df_2018<-df_2018%>% group_by(state) %>% summarise_each(funs(sum))
df_2019<-df_2019 %>% group_by(state) %>% summarise_each(funs(sum))
df_2020<-df_2020 %>% group_by(state) %>% summarise_each(funs(sum))
df_2021<-df_2021 %>% group_by(state) %>% summarise_each(funs(sum))

# Combine casualty dataframe
casualty_df<-full_join(df_2021,df_2020,by="state") %>% 
          full_join(df_2019,by="state") %>% 
          full_join(df_2018,by="state") %>% 
          full_join(df_2017,by="state")
head(casualty_df)

# Convert state name to lower case
casualty_df$state<-tolower(casualty_df$state)

# order state name alphabetically
casualty_df <- casualty_df[order(casualty_df$state),]
casualty_df
```

# Combine unemployment and casualty data and get final data frame
```{r }
df_combined<-full_join(unemp_df,casualty_df,by="state")
df_combined
df_combined[is.na(df_combined)] <- 0
df_combined

# Adding average and total casualty column
df_final<-df_combined %>%
  rowwise() %>%
  mutate(avg_casualty = mean(c_across(total_casualty_2021:total_casualty_2017)),total_casualty=sum(c_across(total_casualty_2021:total_casualty_2017))) %>% select(state,avg_unemp_rate,avg_casualty,total_casualty)
df_final<-data.frame(df_final)
glimpse(df_final)
head(df_final)
```

# Visualize average unemployment rate in states
```{r }
ggplot(unemp_df,aes(x=reorder(state,-avg_unemp_rate), avg_unemp_rate))+geom_bar(stat='identity',width = 0.8,color='red')+coord_flip()
```

# Check state names with highest and lowest average unemployement rate 
```{r }
index1<-which.max(unemp_df$avg_unemp_rate)
unemp_df$state[index1]
unemp_df$avg_unemp_rate[index1]
index2<-which.min(unemp_df$avg_unemp_rate)
unemp_df$state[index2]
unemp_df$avg_unemp_rate[index2]
```

# Visualize total casualty in states 
```{r }
ggplot(df_final,aes(x=reorder(state,-total_casualty), total_casualty))+geom_bar(stat='identity',width = 0.8,color='red')+coord_flip()
```

# Check state names with highest and lowest mass shooting casualty
```{r}
index3<-which.max(df_final$total_casualty)
df_final$state[index3]
df_final$total_casualty[index3]
index4<-which.min(df_final$total_casualty)
df_final$state[index4]
df_final$total_casualty[index4]
```

# Linear regression analysis
plotting the relationship between the average unemployment rate and total casualty in US states for the year between 2017 and 2021 considering average unemployment rate as the predictor (independent variable ).  

```{r }
ggplot(df_final,aes(x=avg_unemp_rate,y=avg_casualty))+geom_point(stat="identity")+geom_smooth(method = "lm")
```

# Find correlation coefficient
```{r }
df_final %>%
summarise(cor(avg_unemp_rate,total_casualty))

```
The coefficient value above indicates a moderate relationship between the variables.

# Find the linear model
By using the lm function to fit the linear model ( a.k.a regression line):

```{r }
linear_model <- lm(total_casualty ~ avg_unemp_rate, data =df_final)
summary(linear_model)

```
With the summary above,the least squares regression line for the linear model: 
yˆ= 131.41 × (avg_unemp_rate)-373.81 

The positive slope line indicates that it's trend is upward. The slope of the line also indicates that if change in average unemployment rate happens by 1, the change in total casualty value will go up by 131.41.
The R-squared value above is very low (0.19) which indicating that the correlation between the two variables is not strong. It is also reflecting that 19% of the variability in the predicting variables can be explained by this model. The p-value,0.00129 is lower than the usual significance level (0.05) which indicating that the average unemployment rate is statistically significant. 

# Model diagnostic 
To assess whether the linear model is reliable, I will check for linearity, nearly normal residual and constant variability of the residuals. 

**Linearity check by the residuals vs. fitted (predicted) plot:
```{r }
ggplot(data = linear_model, aes(x = .fitted, y = .resid)) +
geom_point() +
geom_hline(yintercept = 0, linetype = "dashed") +
xlab("Fitted values") +
ylab("Residuals")
```
From the plot above, it is seen that the residuals are not distributed around 0 with a consistent pattern,
which indicating a  linear trend. So, I can not assume a linear regression model for fitting the data set here.

** Nearly normal residuals check by plotting histogram:
```{r }
ggplot(data = linear_model, aes(x = .resid)) +
geom_histogram(binwidth = 120) +
xlab("Residuals")

```

or by a normal probability plot of the residuals:
```{r }
ggplot(data = linear_model, aes(sample = .resid)) +
stat_qq()
```

It is seen that the histogram is not nearly normally distributed (right skewed). Also, the qq plot is not reflecting a straight line. It has zig-zag and curvature in its shape and errant points. So, the residuals condition for being nearly normal is not met here.

** Constant variability check: From the residuals vs. fitted plot above, it is seen that the spread of the residuals is not roughly equal at each level of the fitted values.So, the constant variance
condition or assumption is not properly met here. 

# Conclusion
The correlation coefficient value (0.44) is suggesting a moderate correlation between the unemployment rate and mass shooting casualty and the p-value (0.00129) is lower than the usual significance level (0.05) is indicating that the unemployment rate is statistically significant.From the linear model diagnostics above, it is seen that the linear regression model was unable to satisfy all the conditions for a strong correlation between the two variables. So, the model will not work properly to predict the relationship between the variables i chose here. Hence, from this analysis it can be said that the unemployment rate in states is not the proper predictor of the mass shooting casualty in US. Rather, it is a relatively a weak predictor of the mass shooting casualties and it is practical. Since, other factors like far-right radicalization, community's change in employment or economic well-being over time have much impact on mass shooting casualties. Therefore, it can be said that the effect of unemployment rate alone is less on mass shooting casualty in USA.  

# References:
Unemployment rate data source: 
https://www.bls.gov/
US mass shooting casualty data source: 
https://www.gunviolencearchive.org/
Introduction to linear regression:
https://fall2022.data606.net/chapters/chapter8/
https://github.com/Raji030/data606_lab8           
             








